---
title: "MALARIA MORTALITY IN SADC REGION"

format:   dashboard
theme: lux
execute:
  echo: false
  warning: false
  message: false

---

The following data presents a summary of reported malaria deaths in the SADC region spanning from 1990 to 2006. This dashboard provides a country-wise summary and trends over the specified period. The dataset is sourced from the gapminder library, focusing on tracking and visualizing changes in malaria-related mortality across Southern African Development Community (SADC) countries during these years..

## Running Code



```{r echo=FALSE,message=F,warning=FALSE}
#Loading required libraries
if(!require("pacman")) install.packages("pacman")
pacman::p_load( here,tidyverse,sf, bslib,  bsicons, rnaturalearth,plotly, countrycode,htmltools, reactable,janitor)
```



```{r echo=FALSE,message=F,warning=FALSE}
#Loading dataset
SADC <- gapminder::gapminder %>%
  filter(continent=="Africa" & country %in% c("Angola", "Botswana","Comoros", "Congo, Dem. Rep.", "Eswatini", "Lesotho", "Madagascar", "Malawi", "Mauritius", "Mozambique", "Namibia", "Seychelles", "South Africa", "Tanzania", "Zambia" , "Zimbabwe")) %>%
mutate(iso3c = countrycode("country", origin = "country.name", destination = "iso3c")) 
  
malaria_df <- read_csv(here("Data/malaria_number_of_deaths_reported.csv")) %>% 
mutate(across(-country, as.numeric)) %>% 
pivot_longer(cols = 2:18,names_to = "period",values_to = "deaths") %>% 
  filter(!is.na(deaths))

SADC_malaria_df <- left_join(SADC,malaria_df) %>%
  mutate(period=as.numeric(period))



countries <- rnaturalearth::ne_countries()

 SADC_malaria_map_df <-inner_join(countries,SADC_malaria_df,by=c("name"="country")) 

```
```{r echo=FALSE,message=F,warning=FALSE}
SADC_malaria_df_final <-  SADC_malaria_df %>% 
  filter(!is.na(deaths)) %>% 
  group_by(period) %>% 
  summarize(deaths=sum(deaths))



 dealth <- SADC_malaria_df_final %>% 
   mutate(total_deaths=sum(deaths)) %>% 
  distinct(total_deaths)

countries_count <- as.data.frame(SADC_malaria_map_df) %>% 
  select(name) %>% 
  distinct()%>% 
   mutate(count=max(NROW(name))) %>% 
  select(-name) %>% 
 mutate(count=max(count)) %>% 
  distinct()


```
# Home

## Row 1 {height=30%} 

```{r echo=FALSE,message=F,warning=FALSE}
value_box(title="Number of Countries Reported",
          value= countries_count,
           theme = value_box_theme(bg="blue"),
          showcase =bs_icon("book",
                  width=4)
          )

```


```{r echo=FALSE,message=F,warning=FALSE}

value_box(title="Malaria Cumulative Deaths",
          value= dealth,
           theme = value_box_theme(bg="Purple"),
          showcase =bs_icon("exclamation",
                  width=4)
          )

```
## Row 2 {height=70%}

```{r echo=FALSE,message=F,warning=FALSE}
  trend_ghp <- ggplot(data=SADC_malaria_df_final,
      aes(x=period,y=deaths,group =1))+
  geom_point(size=3)+
  geom_line()+
    labs(x = "Period", y = "Number of Deaths", title = "Malaria Deaths Over Time") +
  theme_minimal()+
  scale_x_continuous(breaks = seq(min(SADC_malaria_df_final$period), max(SADC_malaria_df_final$period), by = 4))+
    theme(plot.title = element_text(hjust = 0.5))+
    scale_y_continuous(labels = scales::comma_format())
ggplotly(trend_ghp)

```


```{r echo=FALSE,message=F,warning=FALSE}
  map_1 <- SADC_malaria_map_df%>% 
    filter(!is.na(deaths) & period==2006) %>% 
  ggplot()+
  geom_sf(aes(fill=deaths))+
   geom_sf_text(aes(label = geounit), size = 3, color = "black", check_overlap = TRUE) +
  labs(title="Number of Death Reported as of the year 2006")+
  theme(plot.title = element_text(hjust = 0.7))+
   theme_void()


```

```{r echo=FALSE,message=F,warning=FALSE}
ggplotly(map_1)
```

# Download
```{r}
library(htmltools)

htmltools::browsable(
  tagList(
    reactable(
      data = head(SADC_malaria_df),  # Displaying only the first few rows
      searchable = TRUE,  # Optional: Enable searching
      defaultPageSize = 20,  # Optional: Set default page size
      pagination = TRUE,  # Optional: Enable pagination if needed
      columns = list(
        country = colDef(name = "country"),  # Column definitions
        period = colDef(name = "period"),
        deaths     = colDef(name = "deaths    ")
      )
    ),
    tags$button(
      "Download as CSV",
      onclick = "Reactable.downloadDataCSV('SADC_malaria_df_final')"
    )
  )
)
```
